Understanding the Lomb-Scargle Periodogram

### 2017-02-02 PracticalLombScargle.tex
reword CLEAN section

In the general non-uniform case, we measure some signal at a set of $N$ times
which we will denote $\{t_n\}$.
Our observing window function in this case is still a sum of delta functions,
but at non-uniform locations.
Mathematically, the observing window will look like this:

\begin{equation}
  W(t; \{t_n\}) = \sum_{n=1}^{N} \delta(t - t_n)
  \eqlabel{nonuniform-window}
\end{equation}

Applying this window to our true underlying signal $g(t)$, we find an observed
signal of the form:

\begin{eqnarray}
  g_{obs}(t) &=& g(t) W(t; \{t_n\}) \nonumber\\
             &=& \sum_{n=1}^{N} g(t_n)\delta(t - t_n)
  \eqlabel{g-nonuniform}
\end{eqnarray}

Just as in the evenly-sampled case, the Fourier transform of the observed
signal is a convolution of the transforms of the true signal and the window:
\begin{equation}
  \mathcal{F}\{g_{obs}\} = \mathcal{F}\{g\} \ast \mathcal{F}\{W\}
  \eqlabel{g-obs-conv}
\end{equation}
Unlike in the uniform case, the window transform $\mathcal{F}\{W\}$ will
generally {\it not} be a nice clean function; from \eq{FT-def} and
\eq{nonuniform-window}, we can write
\begin{equation}
  \mathcal{F}\{W\}(f) = \sum_{n=1}^{N} e^{-2\pi i f t_n}
  \eqlabel{nonuniform-window-FT}
\end{equation}
For uniformly-spaced observing times $t_n$, the symmetry leads to cancellations
so that the window transform is a clean Dirac comb.
For non-uniform observing times, this symmetry is broken and the window
transform will be much more ``nosiy''.
This can be seen in \fig{random-window}, which shows the Fourier transform of
a non-uniform observing window with an average sampling rate identical to
that in \fig{comb-window-1}, along with its imact on the observed Fourier
transform.

@@

Our discussion of window functions here and in \sect{window-functions}
has highlighted the impact of structure in the survey window on the
resulting observed power spectrum, and the importance of examining that
window power to understand features of the resulting periodogram.
The window power spectrum can be computed directly from the
delta-function representation of the window
function; recalling \eq{nonuniform-window-FT}, the window power spectrum
for observation times $\{t_n\}$ is
\begin{equation}
  \mathcal{P}_W(f;\{t_n\}) = \left|\sum_{n=1}^{N} e^{-2\pi i f t_n}\right|^2
  \eqlabel{window-power-analytic}
\end{equation}
Comparing this to \eq{schuster-periodogram}, we see that this is essentially
the Schuster periodogram for data $g_n=1$ at all times $t_n$;
with that in mind we can compute the Lomb-Scargle window power by computing
a standard Lomb-Scargle periodogram on a series of unit measurements,
making sure to {\it not} pre-center the data or use a floating mean model.
This method is how window power spectra have been computed throughout this
paper.

Recall from \eq{g-nonuniform} that the observed data is a point-wise product
of the underlying function and the survey window:
\begin{equation}
  g_{obs}(t) = g_{true}(t)W_{\{t_n\}}(t),
\end{equation}
and the convolution theorem tells us that the observed Fourier transform is
a convolution of the true transform and the window transform:
\begin{equation}
  \mathcal{F}\{g_{obs}\} = \mathcal{F}\{g_{true}\} \ast \mathcal{F}\{W_{\{t_n\}}\}
\end{equation}
Given this direct relationship, we might hope to be able to invert this
convolution to recover $\mathcal{F}\{g_{true}\}$ directly.
For example, we could write:
\begin{eqnarray}
  \mathcal{F}\{g_{true}\} &=& \mathcal{F}\{g_{obs}/W_{\{t_n\}}\}\nonumber\\
                   &=& \mathcal{F}\{g_{obs}\} \ast \mathcal{F}\{1/W_{\{t_n\}}\}
\end{eqnarray}
Because of the localization of observations, $W_{\{t_n\}}(t)$ is zero for most
values of $t$ and such an inversion is not well-defined.
More precisely, it can be shown that deconvolution in this context is an
ill-posed problem without a unique solution, and so direct deconvolution
is not an option.

There have been a few attempts in the literature to use procedural algorithms
to solve this under-constrained deconvolution problem, perhaps most notably
by adapting the iterative CLEAN algorithm developed for deconvolution in the
context of radio astronomy \citep{Roberts87}.
For cleaning of Lomb-Scargle periodograms, the CLEAN approach is hindered by
three main deficiencies: first and most importantly, the CLEAN algorithm at each
iteration assumes that the highest peak is the location of the primary signal;
this is not always borne out for realistic observations of faint objects where
the cleaning is most necessary (see the discussion in \sect{failure-modes}).
Second, the convolution takes place at the level of the Fourier transform
rather than the PSD; trying to clean a PSD directly ignores important phase
information. Third, the CLEAN algorithm assumes a Schuster-type FFT analysis:
while the Lomb-Scargle periodogram is equivalent to Schuster in the limit of
equally-spaced observations, it is not equivalent in the relevant case of
unequal observations (see \sect{schuster-to-lomb-scargle}),
and so any attempt to apply CLEAN to Lomb-Scargle
analysis directly is fundamentally flawed.

My feeling is that there is room for a more principled approach to the
unconstrained deconvolution problem for periodograms derived from non-uniform
fast Fourier transforms (perhaps through some sort of regularization),
but this does not appear to have been explored anywhere in the astronomy
literature.

>>>>

In the general non-uniform case, we measure some signal at a set of $N$ times
which we will denote $\{t_n\}$.
Our observing window function in this case is still a sum of delta functions,
but at non-uniform locations.
Mathematically, the observing window will look like this:
\begin{equation}
  W_{\{t_n\}}(t) = \sum_{n=1}^{N} \delta(t - t_n)
  \eqlabel{nonuniform-window}
\end{equation}
Applying this window to our true underlying signal $g(t)$, we find an observed
signal of the form:
\begin{eqnarray}
  g_{obs}(t) &=& g(t) W_{\{t_n\}}(t) \nonumber\\
             &=& \sum_{n=1}^{N} g(t_n)\delta(t - t_n)
  \eqlabel{g-nonuniform}
\end{eqnarray}

Just as in the evenly-sampled case, the Fourier transform of the observed
signal is a convolution of the transforms of the true signal and the window:
\begin{equation}
  \mathcal{F}\{g_{obs}\} = \mathcal{F}\{g\} \ast \mathcal{F}\{W_{\{t_n\}}\}
  \eqlabel{g-obs-conv}
\end{equation}
Unlike in the uniform case, the window transform $\mathcal{F}\{W_{\{t_n\}}\}$
will generally {\it not} be a nice clean function; from \eq{FT-def} and
\eq{nonuniform-window}, we can write
\begin{equation}
  \mathcal{F}\{W_{\{t_n\}}\}(f) = \sum_{n=1}^{N} e^{-2\pi i f t_n}
  \eqlabel{nonuniform-window-FT}
\end{equation}
For uniformly-spaced observing times $t_n$, the symmetry leads to cancellations
so that the window transform is a clean Dirac comb.
For non-uniform observing times, this symmetry is broken and the window
transform will be much more ``nosiy''.
This can be seen in \fig{random-window}, which shows the Fourier transform of
a non-uniform observing window with an average sampling rate identical to
that in \fig{comb-window-1}, along with its imact on the observed Fourier
transform.

@@

Our discussion of window functions here and in \sect{window-functions}
has highlighted the impact of structure in the survey window on the
resulting observed power spectrum, and the importance of examining that
window power to understand features of the resulting periodogram.

\subsubsection{Computing the Window Function}
The window power spectrum can be computed directly from the
delta-function representation of the window
function; recalling \eq{nonuniform-window-FT}, the window power spectrum
for observation times $\{t_n\}$ is
\begin{equation}
  \mathcal{P}_W(f;\{t_n\}) = \left|\sum_{n=1}^{N} e^{-2\pi i f t_n}\right|^2
  \eqlabel{window-power-analytic}
\end{equation}
Comparing this to \eq{schuster-periodogram}, we see that this is essentially
the Schuster periodogram for data $g_n=1$ at all times $t_n$;
with that in mind one quick way to compute the window power is to use
a standard Lomb-Scargle periodogram on a series of unit measurements,
making sure to {\it not} pre-center the data or to use a floating mean model.
This method is how window power spectra have been computed throughout this
paper.

\subsubsection{Deconvolution and CLEANing}
With this ability to compute the window function, we might hope to be able to
use it {\it quantitatively} to remove spurious peaks from the observed
power spectrum
Recall from \eq{g-nonuniform} that the observed data is a point-wise product
of the underlying function and the survey window:
\begin{equation}
  g_{obs}(t) = g_{true}(t)W_{\{t_n\}}(t),
\end{equation}
and the convolution theorem tells us that the observed Fourier transform is
a convolution of the true transform and the window transform:
\begin{equation}
  \mathcal{F}\{g_{obs}\} = \mathcal{F}\{g_{true}\} \ast \mathcal{F}\{W_{\{t_n\}}\}
\end{equation}
Given this relationship, we might hope to be able to invert this
convolution to recover $\mathcal{F}\{g_{true}\}$ directly.
For example, we could write:
\begin{eqnarray}
  \mathcal{F}\{g_{true}\} &=& \mathcal{F}\{g_{obs}/W_{\{t_n\}}\}\nonumber\\
                   &=& \mathcal{F}\{g_{obs}\} \ast \mathcal{F}\{1/W_{\{t_n\}}\}
\end{eqnarray}
Because of the localization of observations, $W_{\{t_n\}}(t)$ is zero for most
values of $t$ and so $1/W_{\{t_n\}}(t)$ and its Fourier transform are not well
defined. For this reason, direct deconvolution is not an option in most
cases of interest---in other words, the deconvolution problem is ill-posed
and does not have a unique solution.

There have been a few attempts in the literature to use procedural algorithms
to solve this under-constrained deconvolution problem, perhaps most notably
by adapting the iterative CLEAN algorithm developed for deconvolution in the
context of radio astronomy \citep{Roberts87}.
For cleaning of Lomb-Scargle periodograms, the CLEAN approach is hindered by
three main deficiencies: first and most importantly, the CLEAN algorithm at each
iteration assumes that the highest peak is the location of the primary signal;
this is not always borne out for realistic observations of faint objects where
the cleaning is most necessary (recall the discussion in \sect{failure-modes}).
Second, the convolution takes place at the level of the Fourier transform
rather than the PSD; trying to clean a PSD directly ignores important phase
information. Third, the CLEAN algorithm assumes a Schuster-type FFT analysis:
while the Lomb-Scargle periodogram is equivalent to Schuster in the limit of
equally-spaced observations, it is not equivalent in the relevant case of
unequal observations (see \sect{schuster-to-lomb-scargle}),
and so any attempt to apply CLEAN to Lomb-Scargle analysis directly would
be fundamentally flawed even if it were not ill-posed to begin with.

The latter two issues could be addressed by focusing on the non-uniform
Fourier transform and the Schuster periodogram rather than the Lomb-Scargle
approach, but one would lose the advantageous statistical properties of the
resulting transform, as well as the correspondance with least-squares fitting
that allows useful extensions to models handling a floating mean and
observational noise.
Even in this case, though, the first issue remains quite problematic.
My feeling is that there is room for a more principled approach to the
unconstrained deconvolution problem for periodograms derived from non-uniform
fast Fourier transforms, perhaps through some sort of L1/lasso regularization
that imposes assumptions of sparsity on the true periodogram, but to date this 
does not appear to have been explored anywhere in the literature.
